---
title: "Session 4: Homework 2"
author: "Group 2: Ken Dobson, Fabian Sinn, Carlota Castro Perez, Jieyi Cai, Proud Chaikul, Othman Bensouda"
date: "`r Sys.Date()`"
output:
  html_document:
  theme: flatly
highlight: zenburn
number_sections: yes
toc: yes
toc_float: yes
code_folding: show
---
  
  
```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.5, 
  fig.height=6.5,
  fig.align = "center"
)
```


```{r load-libraries, include=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
library(ggThemeAssist)
library(knitr)
library(infer)
library(kableExtra)
library(wbstats)
library(countrycode)
```


# 1 Climate change and temperature anomalies 

If we wanted to study climate change, we can find data on the *Combined Land-Surface Air and Sea-Surface Water Temperature Anomalies* in the Northern Hemisphere at [NASA's Goddard Institute for Space Studies](https://data.giss.nasa.gov/gistemp). The [tabular data of temperature anomalies can be found here](https://data.giss.nasa.gov/gistemp/tabledata_v3/NH.Ts+dSST.txt)

To define temperature anomalies you need to have a reference, or base, period which NASA clearly states that it is the period between 1951-1980.

```{r weather_data, cache=TRUE, echo=FALSE}

weather <- 
  read_csv("https://data.giss.nasa.gov/gistemp/tabledata_v3/NH.Ts+dSST.csv", 
           skip = 1, 
           na = "***")

```


```{r tidyweather}
#selecting the months only
dates <- weather %>% 
  select(Year:Dec)

#converting dataframe to 'variables'long' format
tidyweather <- dates %>% 
  pivot_longer(col = Jan:Dec,
               names_to= "Month",
               values_to = "delta")

#glimpse(tidyweather)
```


## 1.1 An Increase in Weather Anomaly Throughout the Century

Let us plot the data using a time-series scatter plot, and add a trendline. To do that, we first need to create a new variable called `date` in order to ensure that the `delta` values are plot chronologically. 

```{r scatter_plot}

tidyweather_2 <- tidyweather %>%
  mutate(date = ymd(paste(as.character(Year), Month, "1")),
         month = month(date, label=TRUE),
         year = year(date))

ggplot(tidyweather_2, aes(x= date, y = delta))+
  geom_point()+
  geom_smooth(color="red") +
  theme_bw() +
  labs (
    title = "An Increase in Weather Anomaly Throughout the Century"
  , y = "Temperature Deviation", x = "Year")


```

> The overall trend indicates that there is an increased level of temperature/weather anomaly throughout time. We suspect that the greenhouse effect on earth could be one of the reasons why the delta value could be rising over time.


## 1.2 Every Month has a Similar Weather Anomaly Trend
```{r facet_wrap}

tidyweather_2$Month = factor(tidyweather_2$Month, levels= c("Jan","Feb","Mar","Apr","May", "Jun", "Jul" ,"Aug","Sep","Oct","Nov" ,"Dec"))

ggplot(tidyweather_2, aes(x=date, y = delta))+
  geom_point()+
  geom_smooth(color="red") +
  theme_bw() +
  facet_wrap(~Month) +
  labs (
    title = "Every Month has a Similar Weather Anomaly Trend", y= "Temperature Deviation", x= "Month")

```

> Overall, the effect of increasing temperature has a similar trend throughout the months. However, it seems to be the most pronounced in the first quarter of the year (January-March.) It is also evident that in January, February, March and November there is a large variation in the data points as the data is more spread out.


## 1.3 Present Day Data Shows a Bigger Variation in Temperature\n Anomalies Than Ever Before

```{r,intervals}

comparison <- tidyweather %>% 
  filter(Year>= 1881) %>% 
  #create new variable 'interval', and assign values based on criteria below:
  mutate(interval = case_when(
    Year %in% c(1881:1920) ~ "1881-1920",
    Year %in% c(1921:1950) ~ "1921-1950",
    Year %in% c(1951:1980) ~ "1951-1980",
    Year %in% c(1981:2010) ~ "1981-2010",
    TRUE ~ "2011-present"
  ))

```
```{r, density_plot}

ggplot(comparison, aes(x=delta, fill=interval))+
  geom_density(alpha=0.2) +   #density plot with tranparency set to 20%
  theme_bw() +                #theme
  labs ( x="Temperature Deviation", title = "Present Day Data Shows a Bigger Variation in Temperature\n Anomalies Than Ever Before",
    y     = "Density" 
  )

```

> Approximately, all of the intervals shown display a normal distribution. As displayed, it is observable that the '1981-2010' and '2011-present' interval are the two intervals with the widest base. This implies that they have the most volatile temperature anomalies. We believe that this could be due to global warming. 


## 1.4 Steep Increase of Temperature Deviation in the Past 5 Decades 
```{r averaging}

#creating yearly averages
average_annual_anomaly <- tidyweather %>% 
  group_by(Year) %>%   #grouping data by Year
  
  # creating summaries for mean delta 
  # use `na.rm=TRUE` to eliminate NA (not available) values 
  summarise(annual_average_delta = mean(delta, na.rm=TRUE)) 

#plotting the data:
ggplot(average_annual_anomaly, aes(x=Year, y= annual_average_delta))+
  geom_point()+
  
  #Fit the best fit line, using LOESS method
  geom_smooth() +
  
  #change to theme_bw() to have white background + black frame around plot
  theme_bw() +
  labs (
    title = "Steep Increase of Temperature Deviation in the Past 5 Decades",
    y     = "Average Annual Temperature Deviation"
  )                         


```

> For the past 50 years, temperature deviation value has been increasing at an alarming rate.
